Marginalization of photometric redshift uncertainties

### 2020-07-22 Paper/paper.tex

    We apply this method to the HSC first public data release as an example, and discuss some of the advantages, limitations and possible extensions of this model.

    \da{The paper is structured as follows...}

@@

    \subsection{A new approach}\label{ssec:theory.nz_new}
      Let us consider a discretized description of the $N_i(z)$ as a sum over basis functions $\phi_\alpha(z)$:
      \begin{equation}
        N_i(z)=\sum_i N^\alpha_i\phi_\alpha(z).
      \end{equation}
      For simplicity here we will just treat $N_i(z)$ as a histogram, and therefore in our case the $\phi_\alpha$ are just top-hat functions centered on an equi-spaced grid of redshifts $z_\alpha$. The formalism described below is however valid for any choice of basis functions. In particular, it is likely that future photometric estimators will provide their uncertainty in terms of a several important components and a corresponding covariance matrix.    We describe the full set of redshift distributions by the vector $\vN\equiv\{N^\alpha_i\,\,\,\forall i,\alpha\}$. In what follows, we will assume that $\vN$ has units of galaxy ``counts'' \as{Do we actually need to say this?} \da{This only affects the specific form of the shot-noise covariance Eq. \ref{eq:cov_nz_shot}}. The actual normalization of the $N(z)$ is irrelevant, since it only enters the theory prediction for the power spectrum as a normalized redshift probability distribution (see Eq. \ref{eq:cell_gg_limber}).

@@

      We can now write a likelihood for $\hat{\bf c}$ in which we explicitly marginalize out all the degrees of freedom associated with $\vN$:
      \begin{equation}
        \mathcal{L} \propto  \int \exp\left [-\frac{1}{2} ({\bf c}-\vt)^T C^{-1} ({\bf c}-\vt) -\frac{1}{2} (\vN - \bar{\vN})^T P^{-1} (\vN -\bar{\vN}) \right]  d^N\vN, \label{eq:like2}
      \end{equation}
      Note that in the equation above, $\vt=\vt({\bf q},\vN)$. It is the non-linear dependence of $\vt$ on $\vN$ that prevents us from performing what looks like a trivial integral analytically. In principle, we could solve this problem by making every single element of vector $\vN$ a part of an MCMC chain for ${\bf q}$ and additional hundreds of $\vN$ parameters. This should be feasible using methods that can sample very large parameter spaces, such as Hamiltonian Monte Carlo \cite{1987PhLB..195..216D}. Instead, faced with queue waiting times at NERSC \da{I'm fine leaving this XD, but the referee or journal will complain} we can Taylor expand the theory in $\vN$ to first order around $\bar{\vN}$:
      \begin{equation}
        \vt({\bf q},\vN) \simeq \vt({\bf q},\bar{\vN}) + {\sf T} \left(\vN - \bar{\vN} \right), \label{eq:taylor}
      \end{equation}
      where the matrix ${\sf T}$ is the gradient of ${\bf t}$ with respect to $\vN$
      \begin{equation}
        {\sf T}\equiv \left.\frac{d{\bf t}}{d\vN}\right|_{{\bf q},\bar{\vN}}
        \label{eq:Tmat}
      \end{equation}
      In other words ${\sf T}$ contains a full information about how all the correlation functions or power spectra respond to a small change in $N_i(z)$.

@@

      \subsubsection{$N(z)$ uncertainties}\label{sssec:theory.prior.cv}
        \begin{figure}[ht]
          \centering  
          \includegraphics[width=1.\textwidth]{./corr_CV_0}
          \caption{Correlation matrix associated with the cosmic-variance contribution to the total $N(z)$ covariance for the first tomographic bin of the HSC data.}\label{fig:CV}
        \end{figure}
        The analysis presented in Section \ref{sec:hsc} will make use of a measured redshift distribution estimated from the COSMOS 30-band catalog \cite{2016ApJS..224...24L}, as was done in \cite{1912.08209}. The main sources of uncertainty for this measurement are cosmic variance in the COSMOS field, shot noise and additional uncertainty in the photometric redshifts used to assign galaxies to different bins. Here we will associate each of this sources to a separate covariance that will be added in quadrature to form the final ${\sf C}_N$. \da{Note that there is one fourth source of uncertainty that we do not address here: although the redshifts provided in the COSMOS 30-band catalog have a high accuracy, they are photometric and therefore subject to potential biases. This has been identified as an important source of the relative differences between ongoing cosmic shear surveys \citep{2020A&A...638L...1J}. Calibrating this bias or incorporating it into the $N(z)$ error model is an important task that requires a detailed study of the COSMOS 30-band sample, and that we leave for future work.}

@@

        Finally we estimate the systematic error associated with the choice of photo-$z$ code as follows. Following \cite{1912.08209}, we consider  {\tt DEmP}, {\tt Ephor}, {\tt Ephor\_AB} and {\tt FRANKEN-Z} as some of the best-performing algorithms presented in \cite{2018PASJ...70S...9T}, and constitute a fair representation of the underlying photo-$z$ uncertainties. Note however that all of these codes rely on COSMOS-30 for calibration. We calculate the variance between $N(z)$s estimated from each code by stacking the redshift probability density functions of each source in the HSC data.  We multiply this variance by a factor of 4 and smooth it by convolving it with a Gaussian kernel with standard deviation $\sigma_z=0.1$. The resulting variance vector is added to the diagonal of the pure sample variance covariance matrix. While the choices made when estimating this contribution are somewhat arbitrary, we believe our selection is conservative while maintaining that variance between codes is representative of the underlying systematic uncertainties \da{I think the referee will get hung up on this... Can we say that we've explore the sensitivity to these choices?}.

@@

      These $N(z)$s are shown as blue solid lines in Fig.~\ref{fig:Nzs} \da{We should describe either here or in the previous section what we define as the fiducial $N(z)$s. I guess, from the fact that the original and smoothed versions aren't so different, that the original ones are the sum of all photo-$z$ codes. Is that the case? Should we also show the COSMOS $N(z)$s and corresponding smoothed version? I'd be really curious to see what they look like. Right now it seems like the smoothing does absolutely nothing.}\bh{Resolved?}. The figure also shows the {\sl smoothed} distribution $\bar{N}(z)$s in solid red \da{maybe we should make it dashed so we can actually see the red line}\bh{Did that help?}, defined in Eq. \ref{eq:priorcov}. Our choice of smoothing prior has only a mild effect on the original redshift distribution, mostly removing sharp features at the edges of adjacent $N(z)$ measurements. Also shown in yellow is a realization of the $N(z)$ drawn from a multivariate Gaussian distribution with a mean given by the red line and a covariance given by ${\sl P}$ in Eq. \ref{eq:priorcov} \da{Is this true? If the smoothing prior has any relevance, then it should only allow "smooth" fluctuations. Also, the third bin is kind of suspicious in that all the fluctuations seem to be low.}
      \bh{It is a Gaussian draw from the inverted smoothing prior multiplied by 0.6, see this code line \url{https://github.com/damonge/WeePeeZee/blob/ce50049c6be90ca40b3a903ba782d8bff4b46c78/create_sacc.py##L246}, and \\ $\mathbf{calculate\_smooth\_s\_and\_prior.py}$ for details on the smoothing prior but it should be same as the equation being referenced}.

@@

\section{Conclusions}

>>>>

    The paper is structured as follows. In Section \ref{sec:theory} we derive the basic equations and discuss how the priors on the allowed fluctuations should be determined. In Section \ref{sec:hsc} we  apply this method to the HSC first public data release as an example to demonstrate its applicability in a practical settings. We conclude in Section \ref{sec:conclusiosn} and discuss some of the advantages, limitations and possible extensions of this model.

@@

    \subsection{A new approach}\label{ssec:theory.nz_new}
      Let us consider a discretized description of the $N_i(z)$ as a sum over basis functions $\phi_\alpha(z)$:
      \begin{equation}
        N_i(z)=\sum_i N^\alpha_i\phi_\alpha(z).
      \end{equation}
      For simplicity here we will just treat $N_i(z)$ as a histogram, and therefore in our case the $\phi_\alpha$ are just top-hat functions centered on an equi-spaced grid of redshifts $z_\alpha$. The formalism described below is however valid for any choice of basis functions. In particular, it is likely that future photometric estimators will provide their uncertainty in terms of a several important components and a corresponding covariance matrix.    We describe the full set of redshift distributions by the vector $\vN\equiv\{N^\alpha_i\,\,\,\forall i,\alpha\}$.  Normalization of the $N(z)$ is irrelevant, since it only enters the theory prediction for the power spectrum as a normalized redshift probability distribution (see Eq. \ref{eq:cell_gg_limber}).
      %In what follows, we will assume that $\vN$ has units of galaxy ``counts''
      %\as{Do we actually need to say this?} \da{This only affects the specific form of the shot-noise covariance Eq. \ref{eq:cov_nz_shot}}.

@@

      We can now write a likelihood for $\hat{\bf c}$ in which we explicitly marginalize out all the degrees of freedom associated with $\vN$:
      \begin{equation}
        \mathcal{L} \propto  \int \exp\left [-\frac{1}{2} ({\bf c}-\vt)^T C^{-1} ({\bf c}-\vt) -\frac{1}{2} (\vN - \bar{\vN})^T P^{-1} (\vN -\bar{\vN}) \right]  d^N\vN, \label{eq:like2}
      \end{equation}
      Note that in the equation above, $\vt=\vt({\bf q},\vN)$. It is the non-linear dependence of $\vt$ on $\vN$ that prevents us from performing what looks like a trivial integral analytically. In principle, we could solve this problem by making every single element of vector $\vN$ a part of an MCMC chain for ${\bf q}$ and additional hundreds of $\vN$ parameters. This should be feasible using methods that can sample very large parameter spaces, such as Hamiltonian Monte Carlo \cite{1987PhLB..195..216D}. Instead, faced with queue waiting times at NERSC we Taylor expand the theory in $\vN$ to first order around $\bar{\vN}$:
      \begin{equation}
        \vt({\bf q},\vN) \simeq \vt({\bf q},\bar{\vN}) + {\sf T} \left(\vN - \bar{\vN} \right), \label{eq:taylor}
      \end{equation}
      where the matrix ${\sf T}$ is the gradient of ${\bf t}$ with respect to $\vN$
      \begin{equation}
        {\sf T}\equiv \left.\frac{d{\bf t}}{d\vN}\right|_{{\bf q},\bar{\vN}}
        \label{eq:Tmat}
      \end{equation}
      In other words ${\sf T}$ contains a full information about how all the correlation functions or power spectra respond to a small change in $N_i(z)$.

@@

      \subsubsection{$N(z)$ uncertainties}\label{sssec:theory.prior.cv}
        \begin{figure}[ht]
          \centering  
          \includegraphics[width=1.\textwidth]{./corr_CV_0}
          \caption{Correlation matrix associated with the cosmic-variance contribution to the total $N(z)$ covariance for the first tomographic bin of the HSC data.}\label{fig:CV}
        \end{figure}
        The analysis presented in Section \ref{sec:hsc} will make use of a measured redshift distribution estimated from the COSMOS 30-band catalog \cite{2016ApJS..224...24L}, as was done in \cite{1912.08209}. The main sources of uncertainty for this measurement are cosmic variance in the COSMOS field, shot noise and additional uncertainty in the photometric redshifts used to assign galaxies to different bins. Here we will associate each of this sources to a separate covariance that will be added in quadrature to form the final ${\sf C}_N$. There is fourth source of uncertainty that we do not address here: although the redshifts provided in the COSMOS 30-band catalog have a high accuracy, they are photometric and therefore subject to potential biases. This has been identified as an important source of the relative differences between ongoing cosmic shear surveys \citep{2020A&A...638L...1J}. Calibrating this bias or incorporating it into the $N(z)$ error model is an important task that requires a detailed study of the COSMOS 30-band sample, and that we leave for future work.

@@

        Finally we estimate the systematic error associated with the choice of photo-$z$ code as follows. Following \cite{1912.08209}, we consider  {\tt DEmP}, {\tt Ephor}, {\tt Ephor\_AB} and {\tt FRANKEN-Z} as some of the best-performing algorithms presented in \cite{2018PASJ...70S...9T}, and constitute a fair representation of the underlying photo-$z$ uncertainties. Note however that all of these codes rely on COSMOS-30 for calibration. We calculate the variance between $N(z)$s estimated from each code by stacking the redshift probability density functions of each source in the HSC data.  We multiply this variance by a factor of 4 and smooth it by convolving it with a Gaussian kernel with standard deviation $\sigma_z=0.1$. The resulting variance vector is added to the diagonal of the pure sample variance covariance matrix. We discuss sensitivity to some of these partially subjective choices below.

@@

      These $N(z)$s are shown as blue solid lines in Fig.~\ref{fig:Nzs} \da{We should describe either here or in the previous section what we define as the fiducial $N(z)$s. I guess, from the fact that the original and smoothed versions aren't so different, that the original ones are the sum of all photo-$z$ codes. Is that the case? Should we also show the COSMOS $N(z)$s and corresponding smoothed version? I'd be really curious to see what they look like. Right now it seems like the smoothing does absolutely nothing.}\bh{Resolved?}. The figure also shows the {\sl smoothed} distribution $\bar{N}(z)$s in solid red \da{maybe we should make it dashed so we can actually see the red line}\bh{Did that help?}, defined in Eq. \ref{eq:priorcov}. Our choice of smoothing prior has only a mild effect on the original redshift distribution, mostly removing sharp features at the edges of adjacent $N(z)$ measurements. Also shown in yellow is a realization of the $N(z)$ drawn from a multivariate Gaussian distribution with a mean given by the red line and a covariance given by ${\sl P}$ in Eq. \ref{eq:priorcov} \da{Is this true? If the smoothing prior has any relevance, then it should only allow "smooth" fluctuations. Also, the third bin is kind of suspicious in that all the fluctuations seem to be low.}

%      \bh{It is a Gaussian draw from the inverted smoothing prior multiplied by 0.6, see this code %line\url{https://github.com/damonge/WeePeeZee/blob/ce50049c6be90ca40b3a903ba782d8bff4b46c78/create_sac%c.py##L246}}
%      % , and \\ \texttt{calculate\_smooth\_s\_and\_prior.py} for details on the smoothing prior but %it should be same as the equation being referenced}.

@@

\section{Conclusions}
\label{sec:conclusions}
